Catastrophic Goodhart: regularizing RLHF with KL divergence does not mitigate heavy-tailed reward misspecification

### 2024-10-24 paper/main.tex
fix drake affiliation
add BoN experiment

\author{%
  Thomas Kwa \\
  Independent / FAR Labs \\
  \texttt{kwathomas0@gmail.com} \\
  \And
  Drake Thomas \\
  Anthropic \\
  \texttt{TODO} \\
  \And
  Adrià Garriga-Alonso \\
  FAR AI \\
  \texttt{adria@far.ai} \\
}

@@

\subsection{Heavy-tailed distributions}
\begin{theorem}
\label{thm1}
Given any heavy-tailed reference distribution
\(Q\) over \(\mathbb R\) with mean \(\mu_Q\), and any
\(M, \epsilon > 0\), there is a distribution \(P\) with mean \(\mu_P>M\)
and \(D_{KL}(P \| Q) < \epsilon\).
\end{theorem}

@@

@@

Intuitively, in a heavy-tailed distribution, events with extremely high $x$ are not very rare, so you don’t pay much of a KL penalty to upweight them so they happen about $1/x$ of the time. This is visually illustrated in Figure \ref{fig1}.

\begin{figure}
    \label{fig1}
    \centering
    \includegraphics[width=0.3\linewidth]{goodhart_kl_001.png}
    \includegraphics[width=0.3\linewidth]{goodhart_kl_024.png}
    \includegraphics[width=0.3\linewidth]{goodhart_kl_050.png}
    \caption{As $t \to \infty$, the mean of $X$ (blue bar) grows without bound while KL divergence $D_{KL}(P_t \,\|\, Q)$ (orange bar) goes to 0. The base distribution Q is a Student t-distribution with $df=3$. In this case, high values of X are upweighted to $1/t^{0.8}$; upweighting them to $1/t$ would cause $\mathbb E[X]$ to converge to ~$1$ while KL divergence goes to zero faster.}
\end{figure}

@@

\section{Additional figures}

See figures \ref{fig:pythia-random}, \ref{fig:acg-results}.

@@

\section{Hyperparameters for ACG}

>>>>

\author{%
  Thomas Kwa \\
  Independent / FAR Labs \\
  \texttt{kwathomas0@gmail.com} \\
  \And
  Drake Thomas\thanks{Work was done while not at Anthropic} \\
  Anthropic \\
  \texttt{drake@anthropic.com} \\
  \And
  Adrià Garriga-Alonso \\
  FAR AI \\
  \texttt{adria@far.ai} \\
}

@@

\subsection{KL divergence on heavy- and light-tailed distributions}
\begin{theorem}
\label{thm1}
Given any heavy-tailed reference distribution
\(Q\) over \(\mathbb R\) with mean \(\mu_Q\), and any
\(M, \epsilon > 0\), there is a distribution \(P\) with mean \(\mu_P>M\)
and \(D_{KL}(P \| Q) < \epsilon\).
\end{theorem}

@@

% Reset figure numbering
\renewcommand\thefigure{\thesection.\arabic{figure}}    
\setcounter{figure}{0}

% Reset table numbering
\renewcommand\thetable{\thesection.\arabic{table}}
\setcounter{table}{0}

@@

Intuitively, in a heavy-tailed distribution, events with extremely high $x$ are not very rare, so you don’t pay much of a KL penalty to upweight them so they happen about $1/x$ of the time. This is visually illustrated in Figure \ref{fig_thm1}.

\begin{figure}
    \label{fig_thm1}
    \centering
    \includegraphics[width=0.3\linewidth]{goodhart_kl_001.png}
    \includegraphics[width=0.3\linewidth]{goodhart_kl_024.png}
    \includegraphics[width=0.3\linewidth]{goodhart_kl_050.png}
    \caption{As $t \to \infty$, the mean of $X$ (blue bar) grows without bound while KL divergence $D_{KL}(P_t \,\|\, Q)$ (orange bar) goes to 0. The base distribution Q is a Student t-distribution with $df=3$. In this case, high values of X are upweighted to $1/t^{0.8}$; upweighting them to $1/t$ would cause $\mathbb E[X]$ to converge to ~$1$ while KL divergence goes to zero faster.}
\end{figure}

@@

\setcounter{figure}{0}
\setcounter{table}{0}
\section{Additional experiments and figures}

Figures \ref{fig:pythia-random}, \ref{fig:acg-results} relate to experiments mentioned in the main paper.
In response to reviewer feedback, we added two further experiments to demonstrate the catastrophic Goodhart phenemonon with artificially heavy-tailed reward, one using best-of-N on synthetic distributions and one with PPO on Pythia 1B.

@@

\subsection{Best-of-N experiment}

We created a synthetic experiment by letting reward $U = X + V$, where $X$ and $V$ are independent and sampled from different probability distributions, consistent with our theoretical assumptions. We vary $N$ from 1 to 65536, do 100 trials of taking the best-of-$N$ sample with highest $U$, and note whether $V$ goes towards 0 (overoptimization) or not.

Possible distributions for V are normal and t-distribution with df=10.
Possible distributions for X are normal, t with df=3, t with df=5, lognormal, and Levy. All of these are heavy-tailed except for normal.
V is scaled to a standard deviation of 2 and X has s.d. of 1 (except for the Levy distribution, which has infinite variance), representing that in ordinary regimes most of the variance comes from utility rather than error.

\begin{figure}
    \centering
    \makebox[\textwidth][c]{\includegraphics[width=1.5\linewidth]{images/bon_synthetic.png}}
    \caption{When the error $X$ is normal and thus light-tailed, $V$ increases monotonically with $N$, consistent with our Theorem \ref{thm6}. \\
    However, when both $X$ and $V$ are heavy-tailed, we see results consistent with theorem \ref{thm5}.
    In $5$ of $6$ cases when $X$ is lognormal or student-t, $V$ first increases then starts to decline around $N=10^2$ or $10^3$. When $X$ is (t, df=5) and V is (t, df=10), $V$ instead peaks around $N=10^5$ (but declines afterwards). When X is Levy-distributed, utility never goes significantly above zero (optimization completely fails) because the Levy distribution is too heavy-tailed.}
    \label{fig:pythia-random}
\end{figure}

\subsection{Hyperparameters for ACG}
